---
title: "Analyzing SeaAroundUs Fishing Catch Composition in Large Marine Ecosystems (1950-2019)"
subtitle: "Download and process data"
date: "`r Sys.Date()`"
author: Jorge Mestre Tomás
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: true
    code_folding: hide
    theme: united
    highlight: tango
    fig_width: 7
    fig_height: 6
    fig_caption: true
    df_print: paged
---


```{r setup}
# R Markdown setup
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
run_all <- FALSE

# Custom functions
# Function to get the Sea Around Us API base URL
get_api_base_url <- function() {
  return("https://api.seaaroundus.org/api/v1")
}

# Function to download Sea Around Us data
get_sau_data <- function(
  area = "lme",
  measure = "tonnage",
  dimension = "functionalgroup",
  format = "csv",
  limit = 10,
  sciname = "false",
  region_id = 1,
  attempts = 10,
  out_dir = "./"
) {
  # Construct the download URL based on input parameters
  download_url <- paste0(
    get_api_base_url(), "/", area, "/", measure, "/",
    dimension, "/?format=", format, "&limit=", limit,
    "&sciname=", sciname, "&region_id=", region_id
  )
  
  # Construct the full output file path
  out_file <- file.path(
    out_dir,
    paste0("sau_", area, "_", measure, "_", dimension, "_", region_id, ".zip")
  )
  
  # Attempt to download the file with retry attempts
  cur_attempt <- 1
  while (!file.exists(out_file) & cur_attempt <= attempts) {
    tryCatch(
      download.file(download_url, destfile = out_file, mode = "wb"),
      error = function(e) {
        options(timeout = getOption("timeout") + 1000)  # Increase the timeout on error
      }
    )
    cur_attempt <- cur_attempt + 1
  }
  
  # Check if the file was successfully downloaded
  if (!file.exists(out_file)) {
    stop(paste("Error downloading data from the following URL:", download_url, "\n",
               "Please check your input parameters and ensure that the Sea Around Us API is accessible."))
  }
}

# Factor levels
func_grp_lvls <- c("Pel_SmMd", "Dem_SmMd", "Pel_Lg", "Dem_Lg", 
                   "Sharks_Rays", "Crusts", "Cephs", "Other")
```

# Data Download and Preprocessing

## Required Packages

Data was processed and analyzed using R Project and the following R packages (versions indicated for reproducibility):

```{r packages, eval = run_all}
library(sf) # version 1.0.14
library(tidyverse) # version 2.0.0
```

## Data Source

We obtained catch data from *Sea Around Us* (<https://www.seaaroundus.org/>) for the world's 66 Large Marine Ecosystems (LMEs). We selected data spanning seventy years (1950-2019), including reported data and reconstructed estimates. LMEs represent large oceanic regions near coastlines with generally higher primary productivity than open ocean areas. The catch data, classified into 30 functional groups, was measured in tonnes.

```{r get_sau_data, eval = run_all}
# Download Sea Around Us data
area <- "lme"
measure <- "tonnage"
dimension <- "functionalgroup"
format <- "csv"
limit <- 10
sciname <- "false"
attempts <- 10
out_dir <- "data/raw_data/sau_files"
id_vector <- 1:66

# Loop through each LME to get data
for (id in id_vector){
  get_sau_data(area, measure, dimension, format, limit, sciname, id, attempts, out_dir)
}
```

Catch data (version 50.1) was retrieved using the Sea Around Us API. The data, available in CSV format, was downloaded, unzipped, and subsequently saved as an Rdata file.

> **_NOTE:_** Central Arctic Ocean (ID 64) currently has no catches due to ice cover.

```{r extract_sau_data, eval = run_all}
# Specify the paths to zip files
zip_files <- file.path(
  out_dir,
  paste0("sau_", area, "_", measure, "_", dimension, "_", id_vector, ".zip")
)

# Initialize an empty list to store dataframes
data_list <- list()

# Loop through each zip file
for (i in 1:length(id_vector)) {
  zip_file_path <- zip_files[i]

  # Create a temporary directory for each zip file
  temp_dir <- tempdir()
  
  # Unzip the file to the temporary directory
  unzip(zipfile = zip_file_path, exdir = temp_dir)
  
  # List the files in the temporary directory
  extracted_files <- list.files(temp_dir, full.names = TRUE)
  
  # Find the CSV file
  csv_file <- extracted_files[grep("\\.csv$", extracted_files)]
  
  # Read the CSV file into a data frame
  data <- tryCatch(
    read.csv(csv_file, sep = ",", header = TRUE),
    error = function(x) {
      message(paste("LME ", i, "downloaded empty data"))
      return(NULL)})
  if (is.null(data)) {
    unlink(temp_dir, recursive = TRUE)
    next
  }
  
  # Extract LME name
  lme_name <- gsub(" ", "_", data[1,1])
    
  # Append the dataframe to the list
  data_list[[lme_name]] <- data[, c("area_name", "year", "functional_group", "tonnes")]
  data_list[[lme_name]]$area_id <- i

  # Cleanup: Remove the temporary directory and its contents
  unlink(temp_dir, recursive = TRUE)
}

# Combine dataframes into a single dataframe
sau_original <- do.call(rbind, data_list)

# Save the combined dataframe as an Rdata file
save(sau_original, file = "data/raw_data/sau_original_classification.RData")
```

## Data Preprocessing

We will not work with the Sea Around Us classification of 30 functional groups, but with a more compact classification consisting of 8 functional groups and encompassing the rest. The table below shows the mapping from the original Sea Around Us classification to our new classification. Subsequently, the catch data was aggregated for each functional group in each LME and year.

| **New Classification** | **Sea Around Us Classification**                     |
| ---------------------------------- | ------------------------------------- |
| Pelagics (small and medium) |  Medium bathypelagics (30 - 89 cm)  |
|                                       |  Medium benthopelagics (30 - 89 cm) |
|                                       |  Medium pelagics (30 - 89 cm)       |
|                                       |  Small bathypelagics (<30 cm)       |
|                                       |  Small benthopelagics (<30 cm)      |
|                                       |  Small pelagics (<30 cm)            |
| Demersals (small and medium) |  Medium bathydemersals (30 - 89 cm) |
|                                         |  Medium demersals (30 - 89 cm)      |
|                                         |  Medium reef assoc. fish (30 - 89 cm) |
|                                         |  Small bathydemersals (<30 cm)      |
|                                         |  Small demersals (<30 cm)           |
|                                         |  Small reef assoc. fish (<30 cm)    |
|                                         |  Small to medium flatfishes (<90 cm) |
| Large pelagics             |  Large bathypelagics (>=90 cm)      |
|                                       |  Large benthopelagics (>=90 cm)    |
|                                       |  Large pelagics (>=90 cm)           |
| Large demersals            |  Large bathydemersals (>=90 cm)     |
|                                       |  Large demersals (>=90 cm)          |
|                                       |  Large flatfishes (>=90 cm)        |
|                                       |  Large reef assoc. fish (>=90 cm)  |
| Sharks and rays       |  Large rays (>=90 cm)               |
|                                       |  Large sharks (>=90 cm)             |
|                                       |  Small to medium rays (<90 cm)     |
|                                       |  Small to medium sharks (<90 cm)   |
| Crustaceans                |  Krill                              |
|                                       |  Lobsters, crabs                    |
|                                       |  Shrimps                            |
| Cephalopods                 |  Cephalopods                        |
| Other                              |  Jellyfish                          |
|                                       |  Other demersal invertebrates       |


```{r update_functional_groups, eval = run_all}
# Use our functional group classification
category_matches <- c(
  "Medium bathypelagics (30 - 89 cm)" = "Pel_SmMd",
  "Medium benthopelagics (30 - 89 cm)" = "Pel_SmMd",
  "Medium pelagics (30 - 89 cm)" = "Pel_SmMd",
  "Small bathypelagics (<30 cm)" = "Pel_SmMd",
  "Small benthopelagics (<30 cm)" = "Pel_SmMd",
  "Small pelagics (<30 cm)" = "Pel_SmMd",

  "Medium bathydemersals (30 - 89 cm)" = "Dem_SmMd",
  "Medium demersals (30 - 89 cm)" = "Dem_SmMd",
  "Medium reef assoc. fish (30 - 89 cm)" = "Dem_SmMd",
  "Small bathydemersals (<30 cm)" = "Dem_SmMd",
  "Small demersals (<30 cm)" = "Dem_SmMd",
  "Small reef assoc. fish (<30 cm)" = "Dem_SmMd",
  "Small to medium flatfishes (<90 cm)" = "Dem_SmMd",

  "Large bathypelagics (>=90 cm)" = "Pel_Lg",
  "Large benthopelagics (>=90 cm)" = "Pel_Lg",
  "Large pelagics (>=90 cm)" = "Pel_Lg",

  "Large bathydemersals (>=90 cm)" = "Dem_Lg",
  "Large demersals (>=90 cm)" = "Dem_Lg",
  "Large flatfishes (>=90 cm)" = "Dem_Lg",
  "Large reef assoc. fish (>=90 cm)" = "Dem_Lg",

  "Large rays (>=90 cm)" = "Sharks_Rays",
  "Large sharks (>=90 cm)" = "Sharks_Rays",
  "Small to medium rays (<90 cm)" = "Sharks_Rays",
  "Small to medium sharks (<90 cm)" = "Sharks_Rays",

  "Krill" = "Crusts",
  "Lobsters, crabs" = "Crusts",
  "Shrimps" = "Crusts",

  "Cephalopods" = "Cephs",

  "Jellyfish" = "Other",
  "Other demersal invertebrates" = "Other"
)

# Update categories and aggregate values
sau_update <- sau_original
match_names <- match(sau_update$functional_group, names(category_matches))
sau_update$functional_group <- category_matches[match_names]
sau_update <- sau_update %>% 
  group_by(area_id, area_name, year, functional_group) %>% 
  summarise(tonnes = sum(tonnes))
```

Since we are interested in the temporal and spatial variation of catch composition, we will be working with relative values and compositional data. Compositional data is a type of multivariate data where the portions of the observational vector are non-negative and sum up to a constant. A composition is a vector of $D$ non-negative components $\mathbf{x} = [x_i,...,x_D]$ summing up to a given constant $k$. Then the $D$-part simplex $S^D$ is defined as 

$$
S^D = \Bigl\{ \mathbf{x} = (x_1,...,x_D)' \in R^D | x_i > 0, \sum^D_{i=1} x_i = k \Bigl\}.
$$

In order for all compositions to sum up to the same constant, we applied the closure operation to normalize them to $k = 1$,

$$
C_k(\mathbf{x}) = \left( \frac{k \cdot x_1}{\sum^D_{i=1} x_i}, ...,  \frac{k \cdot x_D}{\sum^D_{i=1} x_i}\right).
$$
This leads to a composition $\mathbf{x}$ with the same number of elements but the vector sum up to 1.

```{r closure_composition, eval = run_all}
# Closure operation
sau_update <- sau_update %>% 
  group_by(area_id, area_name, year) %>% 
  mutate(comp = tonnes / sum(tonnes)) %>% 
  as.data.frame()

save(sau_update, file = "data/raw_data/sau_update_classification.RData")
```

Complete missing values with 0 (essential zeros).

```{r load_compostion, eval = run_all}
# Load data set
load("data/raw_data/sau_update_classification.RData")
load("data/clean_data/lme_name_to_id.RData")

sau_update <- sau_update %>% 
  select(!c(area_name)) %>%
  complete(area_id = 1:66, year, functional_group, fill = list(comp = 0)) %>% 
  left_join(lme_name_to_id, by = "area_id")
sau_update$functional_group <- factor(sau_update$functional_group, levels = func_grp_lvls)
```

# LME polygons processing

We will also load the polygons with the LMEs and compute their centroids for ploting.

```{r lme_polygons_load, results='hide', eval = run_all}
# Get LME polygons
lme_geom <- st_read("data/raw_data/polygons/lme66.kml") %>% 
  select(c(Name, geometry))

# Get LME centroids
x_centroids <- c()
y_centroids <- c()
for (i in 1:66){
  tmp_cent <- lme_geom[i, ] %>% 
    st_make_valid() %>% 
    st_centroid() %>%
    st_geometry() 
    #st_transform(st_crs("+proj=moll"))
  
  x_centroids <- c(x_centroids, tmp_cent[[1]][1])
  y_centroids <- c(y_centroids, tmp_cent[[1]][2])
  
}
lme_geom$x <- x_centroids
lme_geom$y <- y_centroids
lme_geom <- st_wrap_dateline(lme_geom)
  #st_transform(st_crs("+proj=moll"))

lme_geom <- left_join(lme_geom, lme_name_to_id, by = join_by(Name == name_sf))
lme_geom$area_id <- factor(lme_geom$area_id, levels = 1:66)
lme_geom <- sf::st_zm(lme_geom)
```

# Save clean data and polygons

```{r, eval = run_all}
lme_comp <- sau_update
write.table(lme_comp, file = "data/clean_data/lme_comp.csv", sep =  "\t", dec = ".", quote = FALSE, col.names = TRUE, row.names = FALSE)
sf::st_write(lme_geom, "data/clean_data/lme_geom.shp")
```

